How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation.

### 2013-07-30 documents/turtle_paper.tex
almost all citations done.

ROC is a confusion matrix (Table \ref{tab:conf}) statistic that is a descriptor the relationship between the false positive rate (\(FPR\), Eq. \ref{eq:fpr}) of a classification model and the true positive rate (\(TPR\), Eq. \ref{eq:tpr}) of a classification model CITATION. 
The area under the ROC curve (AUC) is a summary statistic of the quality of the classification and varies between 0.5 and 1, with 0.5 indicating a model that classifies no better than random and a value of 1 indicating perfect classification CITATION. AUC can be used as a model selection criterion for classification models and is especially useful in cases where some if not all of the models in question were not fit via maximum likelihood where a criterion such as AICc (see below) or similar can be used CITATION.
\begin{table}
  \centering
  \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
    \cline{3-4} 
    & & \multicolumn{2}{ c |}{Predicted class} \\ \cline{3-4}
    & & 1 & 0 \\ \hline
    \multicolumn{1}{| c |}{\multirow{2}{*}{Actual class}}
    & 1 & TRUE \newline POSITIVE & FALSE \newline NEGATIVE \\ \cline{2-4}
    \multicolumn{1}{| c |}{} & 0 & FALSE \newline POSITIVE & TRUE \newline NEGATIVE \\
    \hline
  \end{tabular}
  \caption{Example confusion matrix. The columns correspond to the predicted class of an observation, while the rows correspond to the actual class of that observation. Depending on the type match between the prediction and reality, four different outcomes are possible: true positive (TP), false negative (FN), false positive (FP), and true negative (TN). These four quantities are used for calculating all confusion matrix statistics.}
  \label{tab:conf}
\end{table}

\begin{equation}
  FPR = \frac{FP}{FP + TN}
  \label{eq:fpr}
\end{equation}
\begin{equation}
  TPR = \frac{TP}{TP + FN}
  \label{eq:tpr}
\end{equation}
%   model selection
%     AICc
For the multinomial logistic regression models, 10 different models were fit each having sequentially more PCs as predictors in order to have models representing different levels of overall amount of shape variation and to estimate how much was necessary and sufficient to best estimate class. The maximum number of PCs allowed as predictors was 10 because of both the large number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately. The final model was that with the lowest AICc \citep{Burnham2002a} AKAIKE AND OTHER CITATION. AICc is a model selection criterion where the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}. Model selection was performed in this manner because the optimal number of PCs to use as predictors was not know \textit{a priori}, and while including all of the PCs of shape would mean that all shape variability would be used to estimate class, this may cause the model to be overfit and not provide an accurate estimate of unsampled plastral variation.

%     recursive feature selection
Random forest models are not fit using maximum likelihood so AICc based model selection was not possible. Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model. PCs were sequentially added as features and the selected model was that with the highest AUC. Similar to the multinomial logistic regression models described above, 10 was the maximum number of PCs that could have been included in the model. After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model. Random forest model parameters were estimated from 1000 subtrees. The recursive feature selection algorithm used in this study was that implemented in the \texttt{caret} package for R \citep{Kuhn2013}.

@@

% multinomial logistic regression model
The AICc best multinomial logistic regression models, for all four classification schemes, all include the first 10 possible PCs as features (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}). The second best models for all classification schemes had the first 9 PCs as features. The \(\Delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis (Table IN SUPPLEMENT?). The first 10 PCs describe 88.6043874205075\% of total variation in plastral shape.

@@

% variation along the most important axes
The results of fitting the final random forest model also include the variable importance for best separating the different classes. The selected random forest model for the best classification scheme had 7 PCs as features, AUC would not increase. Of these 7 features, the first three are illustrated here (Fig. \ref{fig:imp_pc}) in descending order of importance SUPPLEMENT WITH VARIABLE IMPORTANCE INFORMATION?.

@@

\begin{sidewaystable}
  \centering
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Jul 29 07:10:37 2013
{\small
\begin{tabular}{lllllllllllrrrrr}
  \hline
(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
  \hline
+ & + & + & + & + & + & + & + & + & + & + & 22.00 & -245.34 & 537.41 & 0.00 & 0.74 \\ 
  + & + & + & + & + & + & + & + & + & + &  & 20.00 & -248.61 & 539.48 & 2.06 & 0.26 \\ 
  + & + & + & + & + & + & + & + & + &  &  & 18.00 & -255.73 & 549.28 & 11.87 & 0.00 \\ 
  + & + & + & + & + & + & + & + &  &  &  & 16.00 & -267.59 & 568.62 & 31.21 & 0.00 \\ 
  + & + & + & + & + & + &  &  &  &  &  & 12.00 & -283.44 & 591.70 & 54.29 & 0.00 \\ 
  + & + & + & + & + &  &  &  &  &  &  & 10.00 & -286.02 & 592.61 & 55.20 & 0.00 \\ 
  + & + & + & + & + & + & + &  &  &  &  & 14.00 & -282.24 & 593.59 & 56.17 & 0.00 \\ 
  + & + & + & + &  &  &  &  &  &  &  & 8.00 & -307.43 & 631.23 & 93.81 & 0.00 \\ 
  + & + & + &  &  &  &  &  &  &  &  & 6.00 & -340.94 & 694.09 & 156.68 & 0.00 \\ 
  + & + &  &  &  &  &  &  &  &  &  & 4.00 & -345.95 & 700.01 & 162.60 & 0.00 \\ 
   \hline
\end{tabular}
}

@@

\begin{sidewaystable}
  \centering
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Jul 29 07:10:37 2013
{\small
\begin{tabular}{lllllllllllrrrrr}
  \hline
(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
  \hline
+ & + & + & + & + & + & + & + & + & + & + & 22.00 & -242.26 & 531.25 & 0.00 & 0.86 \\ 
  + & + & + & + & + & + & + & + & + & + &  & 20.00 & -246.34 & 534.94 & 3.70 & 0.14 \\ 
  + & + & + & + & + & + & + & + & + &  &  & 18.00 & -252.76 & 543.34 & 12.10 & 0.00 \\ 
  + & + & + & + & + & + & + & + &  &  &  & 16.00 & -263.19 & 559.83 & 28.59 & 0.00 \\ 
  + & + & + & + & + & + &  &  &  &  &  & 12.00 & -275.48 & 575.79 & 44.54 & 0.00 \\ 
  + & + & + & + & + & + & + &  &  &  &  & 14.00 & -273.87 & 576.85 & 45.60 & 0.00 \\ 
  + & + & + & + & + &  &  &  &  &  &  & 10.00 & -280.27 & 581.11 & 49.86 & 0.00 \\ 
  + & + & + & + &  &  &  &  &  &  &  & 8.00 & -303.54 & 623.46 & 92.22 & 0.00 \\ 
  + & + & + &  &  &  &  &  &  &  &  & 6.00 & -342.40 & 697.01 & 165.76 & 0.00 \\ 
  + & + &  &  &  &  &  &  &  &  &  & 4.00 & -348.78 & 705.67 & 174.42 & 0.00 \\ 
   \hline
\end{tabular}
}

@@

\begin{sidewaystable}
  \centering
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Jul 29 07:10:37 2013
{\small
\begin{tabular}{lllllllllllrrrrr}
  \hline
(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
  \hline
+ & + & + & + & + & + & + & + & + & + & + & 33.00 & -297.91 & 668.06 & 0.00 & 0.97 \\ 
  + & + & + & + & + & + & + & + & + & + &  & 30.00 & -305.12 & 675.37 & 7.30 & 0.03 \\ 
  + & + & + & + & + & + & + & + & + &  &  & 27.00 & -314.81 & 687.76 & 19.70 & 0.00 \\ 
  + & + & + & + & + & + & + & + &  &  &  & 24.00 & -329.53 & 710.30 & 42.24 & 0.00 \\ 
  + & + & + & + & + & + & + &  &  &  &  & 21.00 & -342.93 & 730.35 & 62.28 & 0.00 \\ 
  + & + & + & + & + & + &  &  &  &  &  & 18.00 & -353.86 & 745.55 & 77.49 & 0.00 \\ 
  + & + & + & + & + &  &  &  &  &  &  & 15.00 & -357.66 & 746.59 & 78.53 & 0.00 \\ 
  + & + & + & + &  &  &  &  &  &  &  & 12.00 & -385.36 & 795.54 & 127.48 & 0.00 \\ 
  + & + & + &  &  &  &  &  &  &  &  & 9.00 & -424.50 & 867.48 & 199.41 & 0.00 \\ 
  + & + &  &  &  &  &  &  &  &  &  & 6.00 & -442.85 & 897.91 & 229.85 & 0.00 \\ 
   \hline
\end{tabular}
}

@@

\begin{sidewaystable}
  \centering
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Jul 29 07:10:37 2013
{\small
\begin{tabular}{lllllllllllrrrrr}
  \hline
(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
  \hline
+ & + & + & + & + & + & + & + & + & + & + & 33.00 & -253.75 & 579.72 & 0.00 & 1.00 \\ 
  + & + & + & + & + & + & + & + & + & + &  & 30.00 & -267.22 & 599.55 & 19.83 & 0.00 \\ 
  + & + & + & + & + & + & + & + & + &  &  & 27.00 & -275.24 & 608.60 & 28.88 & 0.00 \\ 
  + & + & + & + & + & + & + & + &  &  &  & 24.00 & -302.99 & 657.22 & 77.51 & 0.00 \\ 
  + & + & + & + & + & + & + &  &  &  &  & 21.00 & -307.70 & 659.88 & 80.16 & 0.00 \\ 
  + & + & + & + & + &  &  &  &  &  &  & 15.00 & -327.52 & 686.30 & 106.59 & 0.00 \\ 
  + & + & + & + & + & + &  &  &  &  &  & 18.00 & -324.35 & 686.51 & 106.79 & 0.00 \\ 
  + & + & + & + &  &  &  &  &  &  &  & 12.00 & -350.14 & 725.11 & 145.39 & 0.00 \\ 
  + & + & + &  &  &  &  &  &  &  &  & 9.00 & -390.78 & 800.03 & 220.32 & 0.00 \\ 
  + & + &  &  &  &  &  &  &  &  &  & 6.00 & -405.14 & 822.49 & 242.77 & 0.00 \\ 
   \hline
\end{tabular}
}

>>>>

ROC is a confusion matrix (Table \ref{tab:conf}) statistic that is a descriptor the relationship between the false positive rate (\(FPR\), Eq. \ref{eq:fpr}) of a classification model and the true positive rate (\(TPR\), Eq. \ref{eq:tpr}) of a classification model \citep{Hastie2009}. 
The area under the ROC curve (AUC) is a summary statistic of the quality of the classification and varies between 0.5 and 1, with 0.5 indicating a model that classifies no better than random and a value of 1 indicating perfect classification \citep{Hastie2009}. AUC can be used as a model selection criterion for classification models and is especially useful in cases where some if not all of the models in question were not fit via maximum likelihood where a criterion such as AICc (see below) or similar can be used \citep{Hastie2009}.
\begin{table}
  \centering
  \begin{tabular}[c]{ p{2cm} c | p{2cm} | p{2cm} |}
    \cline{3-4} 
    & & \multicolumn{2}{ c |}{Predicted class} \\ \cline{3-4}
    & & 1 & 0 \\ \hline
    \multicolumn{1}{| c |}{\multirow{2}{*}{Actual class}}
    & 1 & TRUE \newline POSITIVE & FALSE \newline NEGATIVE \\ \cline{2-4}
    \multicolumn{1}{| c |}{} & 0 & FALSE \newline POSITIVE & TRUE \newline NEGATIVE \\
    \hline
  \end{tabular}
  \caption{Example confusion matrix. The columns correspond to the predicted class of an observation, while the rows correspond to the actual class of that observation. Depending on the type match between the prediction and reality, four different outcomes are possible: true positive (TP), false negative (FN), false positive (FP), and true negative (TN). These four quantities are used for calculating all confusion matrix statistics.}
  \label{tab:conf}
\end{table}

\begin{equation}
  FPR = \frac{FP}{FP + TN}
  \label{eq:fpr}
\end{equation}
\begin{equation}
  TPR = \frac{TP}{TP + FN}
  \label{eq:tpr}
\end{equation}
%   model selection
%     AICc
For the multinomial logistic regression models, 10 different models were fit each having sequentially more PCs as predictors in order to have models representing different levels of overall amount of shape variation and to estimate how much was necessary and sufficient to best estimate class. The maximum number of PCs allowed as predictors was 10 because of both the large number of parameters estimated per model and the necessary sample size needed to estimate that many parameters accurately. The final model was that with the lowest AICc \citep{Akaike1974,Hurvich1989,Burnham2002a}. AICc is a model selection criterion where the model with lowest AICc has the fairest variance--bias tradeoff \citep{Burnham2002a}. Model selection was performed in this manner because the optimal number of PCs to use as predictors was not know \textit{a priori}, and while including all of the PCs of shape would mean that all shape variability would be used to estimate class, this may cause the model to be overfit and not provide an accurate estimate of unsampled plastral variation.

%     recursive feature selection
Random forest models are not fit using maximum likelihood so AICc based model selection was not possible. Instead, a recursive feature selection algorithm was used to choose the optimal number of PCs to include based on the AUC of the model. Following the backwards selection algorithm implemented in \texttt{caret} \citep{Kuhn2013}, the maximum number of features included in an initial model, there importance ranked, and the AUC of the model calculated. The lowest ranked feature was then removed, and the AUC of the model recalculated. This was repeated until only one feature, remained. Similar to the multinomial logistic regression models described above, 10 was the maximum number of PCs that could have been included in the model. After each PC was added, 10-fold CV was used to estimate the optimal values of the tuning parameters as well as quantify the uncertainty of each model. Random forest model parameters were estimated from 1000 subtrees. Because PCs were kept in order of importance and not in relation to the amount of variance each PC described, this means that the exact PCs included in each model do not correspond to the PCs in each of the 10 multinomial logistic regressions models.

@@

% multinomial logistic regression model
The AICc best multinomial logistic regression models, for all four classification schemes, all include the first 10 possible PCs as features (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}). The second best models for all classification schemes had the first 9 PCs as features. The \(\Delta\)AICc values between the optimal and second best model range from 2.0639 for the first morphological based classification hypothesis to 19.8349 for the second molecular based classification hypothesis (Tables \ref{tab:mod_sel_1}\ref{tab:mod_sel_2}\ref{tab:mod_sel_3}, and \ref{tab:mod_sel_4}). The first 10 PCs describe 88.6043874205075\% of total variation in plastral shape.

@@

% variation along the most important axes

The results of fitting the final random forest model also include the variable importance for best separating the different classes. The selected random forest model for the best classification scheme had 7 PCs as features, AUC would not increase. Of these 7 features, the first three are illustrated here (Fig. \ref{fig:imp_pc}) in descending order of importance. The PCs included as features in the final random forest model, in descending order of importance, were PCs 3, 8, 6, 1, 4, 2 and 10.

@@

\begin{sidewaystable}
  \centering
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Tue Jul 30 01:18:11 2013
{\small
\begin{tabular}{lllllllllllrrrrr}
  \hline
(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
  \hline
+ & + & + & + & + & + & + & + & + & + & + & 22.00 & -245.34 & 537.41 & 0.00 & 0.74 \\ 
  + & + & + & + & + & + & + & + & + & + &  & 20.00 & -248.61 & 539.48 & 2.06 & 0.26 \\ 
  + & + & + & + & + & + & + & + & + &  &  & 18.00 & -255.73 & 549.28 & 11.87 & 0.00 \\ 
  + & + & + & + & + & + & + & + &  &  &  & 16.00 & -267.59 & 568.62 & 31.21 & 0.00 \\ 
  + & + & + & + & + & + &  &  &  &  &  & 12.00 & -283.44 & 591.70 & 54.29 & 0.00 \\ 
  + & + & + & + & + &  &  &  &  &  &  & 10.00 & -286.02 & 592.61 & 55.20 & 0.00 \\ 
  + & + & + & + & + & + & + &  &  &  &  & 14.00 & -282.24 & 593.59 & 56.17 & 0.00 \\ 
  + & + & + & + &  &  &  &  &  &  &  & 8.00 & -307.43 & 631.23 & 93.81 & 0.00 \\ 
  + & + & + &  &  &  &  &  &  &  &  & 6.00 & -340.94 & 694.09 & 156.68 & 0.00 \\ 
  + & + &  &  &  &  &  &  &  &  &  & 4.00 & -345.95 & 700.01 & 162.60 & 0.00 \\ 
   \hline
\end{tabular}
}

@@

\begin{sidewaystable}
  \centering
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Tue Jul 30 01:18:11 2013
{\small
\begin{tabular}{lllllllllllrrrrr}
  \hline
(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
  \hline
+ & + & + & + & + & + & + & + & + & + & + & 22.00 & -242.26 & 531.25 & 0.00 & 0.86 \\ 
  + & + & + & + & + & + & + & + & + & + &  & 20.00 & -246.34 & 534.94 & 3.70 & 0.14 \\ 
  + & + & + & + & + & + & + & + & + &  &  & 18.00 & -252.76 & 543.34 & 12.10 & 0.00 \\ 
  + & + & + & + & + & + & + & + &  &  &  & 16.00 & -263.19 & 559.83 & 28.59 & 0.00 \\ 
  + & + & + & + & + & + &  &  &  &  &  & 12.00 & -275.48 & 575.79 & 44.54 & 0.00 \\ 
  + & + & + & + & + & + & + &  &  &  &  & 14.00 & -273.87 & 576.85 & 45.60 & 0.00 \\ 
  + & + & + & + & + &  &  &  &  &  &  & 10.00 & -280.27 & 581.11 & 49.86 & 0.00 \\ 
  + & + & + & + &  &  &  &  &  &  &  & 8.00 & -303.54 & 623.46 & 92.22 & 0.00 \\ 
  + & + & + &  &  &  &  &  &  &  &  & 6.00 & -342.40 & 697.01 & 165.76 & 0.00 \\ 
  + & + &  &  &  &  &  &  &  &  &  & 4.00 & -348.78 & 705.67 & 174.42 & 0.00 \\ 
   \hline
\end{tabular}
}

@@

\begin{sidewaystable}
  \centering
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Tue Jul 30 01:18:11 2013
{\small
\begin{tabular}{lllllllllllrrrrr}
  \hline
(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
  \hline
+ & + & + & + & + & + & + & + & + & + & + & 33.00 & -297.91 & 668.06 & 0.00 & 0.97 \\ 
  + & + & + & + & + & + & + & + & + & + &  & 30.00 & -305.12 & 675.37 & 7.30 & 0.03 \\ 
  + & + & + & + & + & + & + & + & + &  &  & 27.00 & -314.81 & 687.76 & 19.70 & 0.00 \\ 
  + & + & + & + & + & + & + & + &  &  &  & 24.00 & -329.53 & 710.30 & 42.24 & 0.00 \\ 
  + & + & + & + & + & + & + &  &  &  &  & 21.00 & -342.93 & 730.35 & 62.28 & 0.00 \\ 
  + & + & + & + & + & + &  &  &  &  &  & 18.00 & -353.86 & 745.55 & 77.49 & 0.00 \\ 
  + & + & + & + & + &  &  &  &  &  &  & 15.00 & -357.66 & 746.59 & 78.53 & 0.00 \\ 
  + & + & + & + &  &  &  &  &  &  &  & 12.00 & -385.36 & 795.54 & 127.48 & 0.00 \\ 
  + & + & + &  &  &  &  &  &  &  &  & 9.00 & -424.50 & 867.48 & 199.41 & 0.00 \\ 
  + & + &  &  &  &  &  &  &  &  &  & 6.00 & -442.85 & 897.91 & 229.85 & 0.00 \\ 
   \hline
\end{tabular}
}

@@

\begin{sidewaystable}
  \centering
% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Tue Jul 30 01:18:11 2013
{\small
\begin{tabular}{lllllllllllrrrrr}
  \hline
(Intercept) & PC1 & PC2 & PC3 & PC4 & PC5 & PC6 & PC7 & PC8 & PC9 & PC10 & df & logLik & AICc & delta & weight \\ 
  \hline
+ & + & + & + & + & + & + & + & + & + & + & 33.00 & -253.75 & 579.72 & 0.00 & 1.00 \\ 
  + & + & + & + & + & + & + & + & + & + &  & 30.00 & -267.22 & 599.55 & 19.83 & 0.00 \\ 
  + & + & + & + & + & + & + & + & + &  &  & 27.00 & -275.24 & 608.60 & 28.88 & 0.00 \\ 
  + & + & + & + & + & + & + & + &  &  &  & 24.00 & -302.99 & 657.22 & 77.51 & 0.00 \\ 
  + & + & + & + & + & + & + &  &  &  &  & 21.00 & -307.70 & 659.88 & 80.16 & 0.00 \\ 
  + & + & + & + & + &  &  &  &  &  &  & 15.00 & -327.52 & 686.30 & 106.59 & 0.00 \\ 
  + & + & + & + & + & + &  &  &  &  &  & 18.00 & -324.35 & 686.51 & 106.79 & 0.00 \\ 
  + & + & + & + &  &  &  &  &  &  &  & 12.00 & -350.14 & 725.11 & 145.39 & 0.00 \\ 
  + & + & + &  &  &  &  &  &  &  &  & 9.00 & -390.78 & 800.03 & 220.32 & 0.00 \\ 
  + & + &  &  &  &  &  &  &  &  &  & 6.00 & -405.14 & 822.49 & 242.77 & 0.00 \\ 
   \hline
\end{tabular}
}
